---
title: "First Draft"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This first section includes the initial inclusion of the People dataset from Sean Lahman's baseball database, and the cleaning of the dataset. 

```{r}
library(dplyr)
library(Lahman)
Lahman_Player_Info <- People %>%
  mutate(finalGame = as.Date(finalGame)) %>% 
  filter(
    birthYear >= 1970,                    
    is.na(finalGame) | finalGame >= as.Date("2015-01-01") 
    
# Ensure that only players in 2015 are included
# I only want a few of these variables, not all of them
    
  ) %>% 
  mutate(nameFull = paste(nameFirst, nameLast)) %>%    
  select(nameFull, weight, height, bats, throws)
```

Now that this part is done, I can read in the Tommy John Surgery list.

```{r}
library(readr)
TJ_Surgery_List <- read_csv("TJ_Surgery_List.csv")
```

This will now only select the variables of choice. 

```{r}
# I do not want all of the variables, so we can get rid of the other ones that will just take up space. 

TJ_Surgery_List <- TJ_Surgery_List %>%
  select(
    "Player",
    "TJ Surgery Date",
    "Team",
    "Level",
    "Position",
    "Throws",
    "Return Date (same level)",
    "Recovery Time (months)"
  )
```


I have to change the dataset so that each of the pitchers with multiple surgeries take up one row, instead of being duplicates. I also want to have it so I can merge these datasets together to make it easier on myself down the road. 

```{r}
TJ_Surgery_List_Numbered <- TJ_Surgery_List %>%
  arrange(Player, `TJ Surgery Date`) %>%        
  group_by(Player) %>%
  mutate(Surgery_Number = row_number()) %>%
  ungroup()


# I need to make sure that each player has one row, no one has multiple rows
library(tidyr)

TJ_Surgery_Wide <- TJ_Surgery_List_Numbered %>%
  pivot_wider(
    id_cols = Player,
    names_from = Surgery_Number,
    values_from = c(
      `TJ Surgery Date`,
      Team,
      Level,
      Position,
      Throws,
      `Return Date (same level)`,
      `Recovery Time (months)`
    ),
    names_glue = "{.value}_Surgery{Surgery_Number}"
  )

# Now I have to merge these two datasets together, TJ and Lahman

Merged_Data <- Lahman_Player_Info %>%
  left_join(TJ_Surgery_Wide, by = c("nameFull" = "Player"))

Merged_Data <- Merged_Data %>%
  mutate(
    Num_Surgeries = rowSums(!is.na(select(., starts_with("TJ Surgery Date_Surgery"))))
  ) %>%
  relocate(Num_Surgeries, .after = throws)

# I want a binomial variable that measures if you had surgery or not
Merged_Data <- Merged_Data %>%
  mutate(
    hadSurgery = ifelse(Num_Surgeries > 0, 1, 0)
  ) %>%
  relocate(hadSurgery, .after = Num_Surgeries)

# I want to add BMI, I think it will be a major indicator. 
Merged_Data <- Merged_Data %>%
  mutate(
    BMI = round((weight / (height^2)) * 703, 2)
  ) %>%
  relocate(BMI, .after = height)
```

Alright, now that the second sort of data cleaning is done, since the NewtForce was done for early results, I can move on to the hard part of Baseball Savant. I have to read in the very large CSV file and clean it a bit. 

```{r}
Baseball_Savant_Metrics <- read.csv("PitchersFullBaseballSavant.csv")

# I need the names to match between the datasets
Baseball_Savant_Metrics <- Baseball_Savant_Metrics %>%
  mutate(
    last_name..first_name = gsub("^\\s+|\\s+$", "",
                                 sub("(.*),\\s*(.*)", "\\2 \\1", last_name..first_name))
  ) %>%
  rename(Player_Name = last_name..first_name)

# I do not need the individual ID if I have the name
Baseball_Savant_Metrics <- Baseball_Savant_Metrics %>%
  select(-player_id)
```

That wasn't as bad as I had originally thought, so now I can work just to clean the NewtForce data now that I have individual metrics like throwing hand, height, weight, and max fastball velocity. 

```{r}
nf_weights_full_data <- read.csv("merged_data.csv")
nf_weights_full_data <- nf_weights_full_data %>%
  mutate(FullName = paste(First.Name, Last.Name)) %>%
  select(-First.Name, -Last.Name) %>%
  relocate(FullName, .before = Date)

```


Now that I have it a bit cleaner, I realize that the names are messed up. Someone really needs to figure out their data collection process. This should not be my problem. 

```{r}
nf_weights_full_data <- nf_weights_full_data %>%
  mutate(
    FullName = FullName %>%
      gsub("Rehab$", "", .) %>%     # remove "Rehab" at end
      gsub(" Rehab$", "", .) %>%    # remove " Rehab" at end
      trimws()                      # remove trailing spaces left behind
  )


# Another important step I came to realize was necessary.
# I need to make sure that each session has an average so that I can analyze with time series

Averaged_Player_Date <- nf_weights_full_data %>%
  filter(FullName %in% c(
    "Ben Hanley",
    "Jaden Bakhit",
    "Beau Chaney",
    "Drew Oerther",
    "Luke Pappano",
    "Boyd Westerfield",
    "Jake Gaerke",
    "Rhys Canan",
    "Tucker Wilburn"
  )) %>%
  group_by(FullName, Date) %>%
  summarise(
    across(
      c(
        `Accel.Impulse..lb.s.`,
        `Accel.Impulse.Score..sec.`,
        `Clawback..sec.`,
        `Player.Velo..mph.`,
        `Stride.Ratio....`,
        `Y.Back.Score..lb.lb.`,
        `Y.Front.Score..lb.lb.`,
        `YZ.Back.Score..lb.lb.`,
        `YZ.Front.Score..lb.lb.`,
        `Z.Transfer..sec.`
      ),
      ~ mean(.x, na.rm = TRUE)
    ),
    .groups = "drop"
  )
```

Once again I need to just run one more check to make sure all of this cleaning worked for the different data sets. One I had to create, the other was inherently created for me to download. 

```{r}
NewtForce_Players_Info <- read.csv("NewtForce_Players_Info.csv")
NewtForce_Players_Info <- NewtForce_Players_Info %>%
  mutate(
    player_name = case_when(
      player_name == "Jaden Bahkit" ~ "Jaden Bakhit",
      player_name == "Rhys Cannon"  ~ "Rhys Canan",
      TRUE ~ player_name
    )
  )
```

I finally can work to merge the two together now that they should, in theory, line up and merge together. 

```{r}
Merged_NewtForce <- Averaged_Player_Date %>%
  left_join(NewtForce_Players_Info,
            by = c("FullName" = "player_name"))

Merged_NewtForce <- Merged_NewtForce %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%Y")) %>%
  arrange(FullName, Date)
```

Now I got another idea to create a data set that takes the TJ surgery list and includes only pitchers that have had surgery. 

```{r}
Merged_Data_Surgery_Yes_Pitchers <- Merged_Data %>%
  filter(
    hadSurgery == 1,
    Position_Surgery1 == "P"
  )
```


## Results

Of course, I found one more thing. Now I am merging the Surgery info into the Baseball Savant metrics. I haven't figured out how I want to do this, but I have a feeling I can find some overlap. 

```{r}
Pitcher_Surgery_Info <- Merged_Data %>%
  select(nameFull, BMI, throws, Num_Surgeries, hadSurgery)

Baseball_Savant_BMI_Included <- Baseball_Savant_Metrics %>%
  left_join(
    Pitcher_Surgery_Info,
    by = c("Player_Name" = "nameFull")
  )
```

Yet another snag in the data. I have to get rid of the accents and periods in the names like AJ, or the Dominican names like Ramirez. 

```{r}
library(stringi)

Baseball_Savant_Metrics <- Baseball_Savant_Metrics %>%
  mutate(
    Player_Name_clean = Player_Name %>%
      stri_trans_general("Latin-ASCII") %>%   # remove accents
      gsub("\\.", "", .) %>%                  # remove ALL periods
      trimws()                                # remove whitespace
  )

Merged_Data <- Merged_Data %>%
  mutate(
    nameFull_clean = nameFull %>%
      stri_trans_general("Latin-ASCII") %>%   # remove accents
      gsub("\\.", "", .) %>%                  # remove ALL periods
      trimws()
  )


Pitcher_Surgery_Info <- Merged_Data %>%
  select(nameFull_clean, BMI, throws, Num_Surgeries, hadSurgery)


Baseball_Savant_BMI_Included <- Baseball_Savant_Metrics %>%
  left_join(
    Pitcher_Surgery_Info,
    by = c("Player_Name_clean" = "nameFull_clean")
  )

Baseball_Savant_BMI_Included <- Baseball_Savant_BMI_Included %>%
  filter(!is.na(BMI))

Baseball_Savant_BMI_Included <- Baseball_Savant_BMI_Included %>%
  filter(year != 2025)

table(is.na(Baseball_Savant_BMI_Included$BMI))
```

I am now ready to make the three scatterplot comparison visuals. I hope these give me something I can use and aren't all worthless. This data cleaning was a lot. My commenting has been awful. 

```{r}
library(dplyr)
library(ggplot2)

# Correlation: BMI and Velo
nf_cor <- cor.test(
  NewtForce_Players_Info$BMI,
  NewtForce_Players_Info$top_fb_velocity,
  method = "pearson"
)

print(nf_cor)

# Linear model
nf_model <- lm(top_fb_velocity ~ BMI, data = NewtForce_Players_Info)
summary(nf_model)


Savant_LatestYear <- Baseball_Savant_BMI_Included %>%
  filter(!is.na(BMI), !is.na(fastball_avg_speed)) %>%          
  group_by(Player_Name) %>%
  filter(year == max(year, na.rm = TRUE)) %>%                  
  summarise(
    BMI = first(BMI),                                          
    fb_velocity = max(fastball_avg_speed, na.rm = TRUE),       
    hadSurgery = first(hadSurgery),                            
    .groups = "drop"
  ) %>%
  filter(is.finite(fb_velocity))                               

# Correlation
bs_cor <- cor.test(
  Savant_LatestYear$BMI,
  Savant_LatestYear$fb_velocity,
  method = "pearson"
)

print(bs_cor)

# Linear model
bs_model <- lm(fb_velocity ~ BMI, data = Savant_LatestYear)
summary(bs_model)

global_x_range <- range(
  NewtForce_Players_Info$BMI,
  Savant_LatestYear$BMI,
  na.rm = TRUE
)

global_y_range <- range(
  NewtForce_Players_Info$top_fb_velocity,
  Savant_LatestYear$fb_velocity,
  na.rm = TRUE
)


Savant_LatestYear_surg1 <- Savant_LatestYear %>%
  filter(hadSurgery == 1)

Savant_LatestYear_surg0 <- Savant_LatestYear %>%
  filter(hadSurgery == 0)

nf_plot <- ggplot(NewtForce_Players_Info, aes(x = BMI, y = top_fb_velocity)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  coord_cartesian(xlim = global_x_range, ylim = global_y_range) +
  labs(
    title = "BMI vs Top Fastball Velocity (NewtForce - All Pitchers)",
    x = "BMI",
    y = "Top Fastball Velocity (mph)"
  ) +
  theme_minimal()

bs_plot_surg1 <- ggplot(Savant_LatestYear_surg1, aes(x = BMI, y = fb_velocity)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  coord_cartesian(xlim = global_x_range, ylim = global_y_range) +
  labs(
    title = "BMI vs Top Fastball Velocity (Savant - Had Surgery)",
    x = "BMI",
    y = "Fastball Velocity (mph)"
  ) +
  theme_minimal()

bs_plot_surg0 <- ggplot(Savant_LatestYear_surg0, aes(x = BMI, y = fb_velocity)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  coord_cartesian(xlim = global_x_range, ylim = global_y_range) +
  labs(
    title = "BMI vs Top Fastball Velocity (Savant - No Surgery)",
    x = "BMI",
    y = "Fastball Velocity (mph)"
  ) +
  theme_minimal()

nf_plot
bs_plot_surg1
bs_plot_surg0

```


Perfect, now that I have the scatterplots I can work to compare the groups statistically. I don't think it'll do any good to use the NF one. I think that one is pointless. 
```{r}

Savant_group_surg1 <- Savant_LatestYear_surg1 %>%
  transmute(
    group = "Savant_Surgery",
    BMI,
    velo = fb_velocity
  )

Savant_group_surg0 <- Savant_LatestYear_surg0 %>%
  transmute(
    group = "Savant_NoSurgery",
    BMI,
    velo = fb_velocity
  )

Combined_groups <- bind_rows(NF_group, Savant_group_surg1, Savant_group_surg0)

interaction_model <- lm(velo ~ BMI * group, data = Combined_groups)
summary(interaction_model)

```
Good. 

Now I can use the merged NF data to expand on my early results time series. I want to do it for all nine guys and make a visual that I can show the comparison of the guys across their sessions. 

```{r}

Merged_NewtForce <- Merged_NewtForce %>%
  mutate(Date = as.Date(Date))


players_of_interest <- c(
  "Ben Hanley",
  "Jaden Bakhit",
  "Beau Chaney",
  "Drew Oerther",
  "Luke Pappano",
  "Boyd Westerfield",
  "Jake Gaerke",
  "Rhys Canan",
  "Tucker Wilburn"
)


Accel_Time_Data <- Merged_NewtForce %>%
  filter(FullName %in% players_of_interest) %>%
  group_by(FullName, Date) %>%
  summarise(
    Accel_Impulse_lb_s = mean(`Accel.Impulse..lb.s.`, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    FullName = factor(FullName, levels = players_of_interest)
  )

accel_plot <- ggplot(Accel_Time_Data, aes(x = Date, y = Accel_Impulse_lb_s)) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~ FullName, ncol = 2) + 
  labs(
    title = "Average Peak Force Across NewtForce Sessions",
    x = "Session Date",
    y = "Accel Impulse (lb·s)"
  ) +
  scale_x_date(date_labels = "%b %Y") +  
  theme_bw(base_size = 14) +            
  theme(
    strip.text = element_text(face = "bold"),       
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.spacing = unit(0.7, "lines")              
  )

accel_plot

```
Let's go!

Now time for a better heat map. First one for early results did not get graded well so want to slim it down and make sure it is readable and usable. 

```{r}
library(tibble)


corr_data <- Baseball_Savant_BMI_Included %>%
  select(
    hadSurgery,
    Num_Surgeries,
    p_formatted_ip,
    n,
    arm_angle,
    fastball_avg_speed,
    breaking_avg_speed,
    offspeed_avg_speed
  ) %>%
  select(where(is.numeric))   

corr_matrix <- cor(corr_data, use = "pairwise.complete.obs")


corr_df <- as.data.frame(corr_matrix) %>%
  rownames_to_column(var = "Var1") %>%
  pivot_longer(
    cols = -Var1,
    names_to = "Var2",
    values_to = "Correlation"
  )

heatmap_corr <- ggplot(corr_df, aes(x = Var2, y = Var1, fill = Correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", Correlation)), size = 3) +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white",
    midpoint = 0,
    limits = c(-1, 1),
    name = "r"
  ) +
  labs(
    title = "Correlation Heatmap: Surgery, IP, Pitch Count, Arm Angle, and Pitch Velocities",
    x = "",
    y = ""
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    panel.grid = element_blank(),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

heatmap_corr

```

Much better than early results.

The bitter end. Now I want to make a LASSO model that predicts injury. Based on the correlation... I feel like this might actually end up being decently strong. I really hope that this is a significant analysis. I think I will go RF or just a log regression if not... LASSO for now. 


```{r}
library(glmnet)
library(caret)
library(pROC)
library(ggthemes)

df <- Baseball_Savant_BMI_Included %>%
  transmute(
    hadSurgery = hadSurgery,
    BMI = BMI,
    n = n,
    p_formatted_ip = as.numeric(p_formatted_ip),
    arm_angle = arm_angle,
    fastball_avg_speed = fastball_avg_speed,
    breaking_avg_speed = breaking_avg_speed,
    offspeed_avg_speed = offspeed_avg_speed
  ) %>%
  filter(!is.na(hadSurgery))

for (col in names(df)) {
  if (col != "hadSurgery") {
    df[[col]][is.na(df[[col]])] <- median(df[[col]], na.rm = TRUE)
  }
}

df$hadSurgery <- as.numeric(df$hadSurgery)

set.seed(123)
train_idx <- createDataPartition(df$hadSurgery, p = 0.8, list = FALSE)
train_df <- df[train_idx, ]
test_df  <- df[-train_idx, ]

x_train <- model.matrix(hadSurgery ~ ., data = train_df)[, -1]
y_train <- train_df$hadSurgery

x_test <- model.matrix(hadSurgery ~ ., data = test_df)[, -1]
y_test <- test_df$hadSurgery

set.seed(123)
cv_lasso <- cv.glmnet(
  x_train, y_train,
  family = "binomial",
  alpha = 1,
  nfolds = 10,
  type.measure = "auc"
)

lambda_min <- cv_lasso$lambda.min

lasso_model <- glmnet(
  x_train, y_train,
  family = "binomial",
  alpha = 1,
  lambda = lambda_min
)

coef_lasso <- as.matrix(coef(lasso_model))

coef_df <- data.frame(
  variable = rownames(coef_lasso),
  coefficient = coef_lasso[, 1],
  row.names = NULL
) %>%
  filter(variable != "(Intercept)") %>%
  arrange(desc(abs(coefficient)))

prob_test <- predict(lasso_model, newx = x_test, type = "response")[, 1]
roc_obj <- roc(response = y_test, predictor = prob_test)
auc_val <- as.numeric(auc(roc_obj))


coef_df <- coef_df %>%
  mutate(
    coefficient = round(coefficient, 3),
    AUC = round(auc_val, 3)
  )

coef_df  

roc_plot <- ggplot(
  data.frame(
    fpr = rev(roc_obj$specificities),
    tpr = rev(roc_obj$sensitivities)
  ),
  aes(x = 1 - fpr, y = tpr)
) +
  geom_line(size = 1.2, color = "steelblue") +
  geom_abline(linetype = "dashed") +
  labs(
    title = paste0("ROC Curve - LASSO Logistic Regression (AUC = ", round(auc_val, 3), ")"),
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +
  theme_wsj(base_size = 14)

roc_plot
ggsave("lasso_roc_curve.png", roc_plot, width = 6, height = 6, dpi = 300)

coef_plot <- coef_df %>%
  ggplot(aes(x = reorder(variable, coefficient), y = coefficient)) +
  geom_bar(stat = "identity", fill = "firebrick", alpha = 0.8) +
  coord_flip() +
  labs(
    title = "LASSO Coefficients\n(Predicting Tommy John Surgery)",
    x = "",
    y = "Coefficient (Standardized)"
  ) +
  theme_wsj(base_size = 14)

coef_plot
ggsave("lasso_coefficients.png", coef_plot, width = 7, height = 5, dpi = 300)

library(gt)

# Adding a table so I can read it better
simple_table <- final_table %>%
  select(variable, coefficient, AUC) %>%
  mutate(
    coefficient = round(coefficient, 3),
    AUC = round(AUC, 3)
  ) %>%
  gt() %>%
  tab_header(
    title = "LASSO Coefficients and Model AUC"
  ) %>%
  fmt_number(columns = c(coefficient, AUC), decimals = 3) %>%
  tab_options(
    table.border.top.style = "solid",
    table.border.bottom.style = "solid",
    table_body.border.top.style = "solid",
    table_body.border.bottom.style = "solid",
    column_labels.border.bottom.style = "solid",
    data_row.padding = px(4)
  ) %>%
  tab_style(
    style = cell_borders(sides = "all", color = "black", weight = px(1)),
    locations = cells_body()
  ) %>%
  tab_style(
    style = cell_borders(sides = "all", color = "black", weight = px(1)),
    locations = cells_column_labels()
  )

simple_table
```
Let's go! First draft bang!